Fitness effects of mutations to SARS-CoV-2 proteins

### 2023-01-27 paper/paper.tex
add Neher fitness calculations

The ratio of actual to expected counts for each mutation is related to its effect on viral fitness.
The intuition is straightforward: mutations arise at all sites during viral replication, but viruses with deleterious mutations are less likely to transmit and be observed in sequencing of human SARS-CoV-2.
Therefore, the ratio of actual to expected counts will be one for neutral mutations, and less than one for deleterious mutations.
In the Methods, we show that under plausible assumptions about SARS-CoV-2 evolution and sampling intensity (fraction of viruses sequenced), the fitness effects of deleterious mutations scales roughly logarithmically with the ratio of actual to expected counts for mutations with effects greater than a few percent.
We therefore estimate the effect of each mutation as the logarithm of this ratio after summing counts for all nucleotides that encode the relevant amino-acid.
The statistical noise in the count ratios is greater for mutations with fewer expected counts: the figures in this paper show mutations with $\ge10$ expected counts unless otherwise noted, with legends linking to interactive plots that enable adjustment of this threshold.

@@

\subsection{Derivation of relationship between actual to expected count ratio and viral fitness}
\jdbcomment{Jesse will add this from Richard's notes.}

@@

\onecolumn
\renewcommand{\thepage}{S\arabic{page}}
\setcounter{page}{1}
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}

@@

\section{Supplementary Material}

>>>>

The ratio of actual to expected counts for each mutation is related to its effect on viral fitness.
The intuition is straightforward: mutations arise at all sites during viral replication, but viruses with deleterious mutations are less likely to transmit and be observed in sequencing of human SARS-CoV-2.
Therefore, the ratio of actual to expected counts will be one for neutral mutations, and less than one for deleterious mutations.
In the Methods and Appendix, we show that under plausible assumptions about SARS-CoV-2 evolution and sampling intensity (fraction of viruses sequenced), the fitness cost of a deleterious mutation scales roughly inversely with the ratio of actual to expected counts for mutations with costs greater than a few percent.
A key result from this derivation is the dependence on sampling intensity: if all human SARS-CoV-2 were sequenced then our approach would only be sensitive to highly deleterious mutations, but since the actual sampling intensity is $\sim$0.1--1\% the approach can detect mutations with more subtle effects that transmit with reduced efficiency.
We quantify the effect of each mutation as the logarithm of the ratio of actual to expected counts after summing counts for all nucleotides that encode the relevant amino-acid.
The statistical noise is greater for mutations with fewer expected counts: the figures in this paper show mutations with $\ge10$ expected counts unless otherwise noted, with legends linking to interactive plots that enable adjustment of this threshold.

@@

\subsection{Derivation of relationship between actual to expected count ratio and viral fitness}
The ratio of actual to expected counts that we calculate in this paper is related to the probability that we observe a viral lineage containing an occurrence of a specific mutation among sequenced human SARS-CoV-2.
This probability depends on three factors: the fitness effect of the mutation, the fraction of all SARS-CoV-2 viruses that are sequenced (sampling intensity), and the growth dynamics of the viral population.
In the supplementary appendix, we derive the approximate relationship between this probability as a function of the fitness cost $s$ and sampling intensity $\epsilon$ for deleterious mutations for both a constant and exponentially growing viral population.

We show that for a constant viral population size, the probability of observing a lineage containing a deleterious mutation with cost $s$ is roughly $\frac{\epsilon}{s + \epsilon}$ when $s^2 > \epsilon$, and more weakly dependent on $s$ for smaller fitness costs (when $s^2 < \epsilon$).
The intuitive explanation is that the average size of a mutant lineage with fitness cost $s$ is $1/s$ and we basically ask whether we sample the lineage before it disappears.
If we sample more intensely (larger $\epsilon$), whether a lineage gets sampled depends primarily on the stochastic dynamics and little on the fitness effect.
With a typical sampling intensity for SARS-CoV-2 between 1/1000 and 1/100, this means our approach is sensitive to fitness effects larger than a few percent; mutations with fitness costs smaller than that will not show an appreciable difference from neutral mutations in their ratio of actual to expected accounts.

In an exponentially growing population, the probability of observing a mutant lineage with fitness cost $s$ again scales as $\sim \frac{\epsilon}{\epsilon + s}$ if $sT>1$, where $T$ is the time over which the variant has expanded.
If $T$ is $\sim$ months, that is 20 generations, which again corresponds to $s$ of at least a few percent for $sT > 1$. 
For mutations with smaller fitness costs, the dependence scales more as $\sim \epsilon\left(1-sT\right)$.

Overall, these calculations indicate that for multiple different growth dynamics of the viral population, the ratio of expected to actual counts will scale inversely with the fitness cost of deleterious mutations for mutations with costs that exceed a few percent.
Note that the approach we use in this paper does not account for variation in sampling intensity across space or time, does not attempt to adjust for changes in viral growth dynamics over time, uses the heuristic formula of calculating the effect as the log ratio of counts, and applies this same formula to all mutations regardless of whether they are deleterious, neutral, or beneficial.
A more complete derivation might try to calculate the fitness effects from the full distribution of lineage sizes more rigorously and incorporate information about the sampling intensity and viral growth dynamics.
However, such a derivation (if possible at all) is beyond the scope of this study, and we also note that good empirical data is generally lacking to precisely account for sampling intensity and viral growth dynamics over the full span of time and space from which the sequences we analyze are drawn.
The key point of the derivations for our current study is simply that our approach should be sensitive to detecting the effects of mutations with fitness costs greater than a few percent.

@@

\newgeometry{left=1in, right=1in, top=48pt, bottom=60pt}
\onecolumn
\renewcommand{\thepage}{S\arabic{page}}
\setcounter{page}{1}
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}

@@

\section{Supplementary appendix deriving relationship between fitness cost and ratio of expected to actual counts}

With millions of SARS-CoV-2 sequences shared publicly, almost all mutations that are tolerated by the virus are observed dozens to hundreds of times.
Where on the tree and how often on the tree we observe specific mutations has information about the effects of these mutations on viral spread.
The mutation rate depends on the nucleotides involved and possibly on the sequence context and other viral determinants, but for the purpose of this derivation, we will assume the neutral rate $\mu$ is known.
If the mutation is neutral, the total number of times the mutation is observed on the tree is $\mu T$, where $T$ is the total length of the tree (assuming that the mutation never reached high frequency which is true for almost all mutations, particularly when mutations are counted on a per-clade basis relative too the clade founder as done above).

If a mutation reduces fitness, the lineages descending from branches on which this mutation happened will spread more slowly than those without this mutation.
As a result, the down-stream subclades are smaller and more short lived, which in turn means that they will be less likely to be sampled and represented in the tree.
To infer a mutation's effect on fitness, we need to calculate how the probability of observation depends on the this fitness effect.

For a mutation to be represented in the tree, one of its descendants has to be sampled and sequenced.
If the total number of descendants is $w$ and the sampling fraction is $\epsilon$, the probability that the mutation is present in the tree is
\begin{equation}
    P = 1 - e^{-w\epsilon}
\end{equation}
$W$ is a random number that depends on the realization of the transmission process, which is commonly modeled by a branching process with birth rate $b$ and death rate $d$.
The death rate here corresponds to clearing an infection, the birth rate to onward transmission. The latter is affected by the fitness cost of the mutation.

To obtain insight how the probability of observing a lineage depends on parameters, we calculate the probability $p(w, T|t)$ that a lineage had an integrated size $w = \int_t^T k(t') \; dt'$, where $t$ is the birth time of the lineage, $T$ is the current time, and $k(t')$ is the size of the lineage at time $t'$.
To calculate $p(w,T|k)$, we generalize it slightly to $p(W, T|k,t)$, where $k$ is the number of individuals at the start time $t$. This quantity obeys the following ``first-step'' equation:
\begin{equation}
    -(\partial_t - k \partial_w) p(w,T|k,t) = -k(b+d) p(w,T|k,t) + k b p(w,T|k+1,t) + k d p(w,T| k-1, t)
\end{equation}
We will solve for the Laplace transform $\hat{p}(z,T|k,t) = \int_0^\infty dw\; e^{-wz} p(w,T|k,t) = \hat{p}^k(z,T|1,t)$.
Using the identity for the derivative of the Laplace transform
\begin{equation}
    \begin{split}
        \int_0^\infty e^{-wz} \partial_w p  \; dw& = [e^{-wz} p]_0^\infty - \int_0^\infty  p \partial_w e^{-wz} \; dw = 0 + z \int_0^\infty  p e^{-wz} = z \hat{p}
    \end{split}
\end{equation}
and using $k=1$, we have
\begin{equation}
    -\partial_t \hat{p}(z,T|t) = -(b+d+z) \hat{p}(z,T|t) + b \hat{p}^2(z,T|t) + d
\end{equation}
This simplifies further to if we substitute $\phi(z,T|t) = 1 - \hat{p}(z,T|t)$.
\begin{equation}
    \begin{split}
        \partial_t \phi(z,T|t) & = -(b+d+z) (1-\phi(z,T|t)) + b (1-\phi(z,T|t))^2 + d \\
         & = -z - (b-d-z) \phi(z,T|t) + b \phi(z,T|t)^2
    \end{split}
\end{equation}
where it is important to note that the derivative is with respect to the first time point and the interval $T-t$ is shrinking with increasing $t$.

\subsection*{Constant birth and death rate}
If the fitness effect of the mutation in question is detrimental and the overall population is constant (background $b_0=d_0$), all mutant lineages will eventually die out and we can consider large $T-t$ and $\partial_t \phi(z,T|t)$.
Further define $b=b_0 - s$ and $d=d_0$ where $s$ is the fitness cost of the mutation (so larger values indicate a greater fitness cost).
The steady state generating function is then
\begin{equation}
    0 = -z - (b-d-z) \phi(z) + b \phi(z)^2
\end{equation}
with solution
\begin{equation}
    \begin{split}
        \phi(z) & = -\frac{s + z}{2(b_0 - s)} \pm \frac{\sqrt{(s+z)^2 + 4z(b_0-s)}}{2(b_0-s)} \\
        &\approx -\frac{s + z}{2b_0} \pm \frac{\sqrt{(s+z)^2 + 4z b_0}}{2b_0} \\
        & \approx \begin{cases}
            \frac{z}{s+z} & (s+z)^2 \gg 4zb_0 \\
            \sqrt{\frac{z}{b_0}}\left(1+\frac{(s+z)^2}{8zb_0}\right)-\frac{s + z}{2b_0}  & (s+z)^2 \ll 4zb_0 \\
        \end{cases}
    \end{split}
\end{equation}
Since $\phi(z) = 1 - \int e^{-wz}p(w) \; dw$, $\phi(\epsilon)$ is exactly the probability that a lineage is sampled when the entire population is sampled at rate $\epsilon$.
We thus expect two regimes: if the square of the fitness cost exceeds the sampling intensity (typically at 1\% or less), the probability of sampling a lineage is essentially inversely proportional to the fitness cost.
The sampling probability of lineages with smaller costs effects depends less strongly on $s$.
Their sampling mostly comes down to stochasticity independent of the fitness cost.

\subsection*{Growing populations}
In many scenarios relevant for lineages that arise during a viral outbreak, the background population isn't constant but is undergoing a rapid exponential expansion.
The background birth rate $b_0$ is bigger than $d_0$ in this case.
Since the population is growing, deleterious mutations can increase in frequency deterministically and we can not send the $t$ to infinity as before.
Instead, we need to integrate
\begin{equation}
    \begin{split}
        \partial_t \phi(z,T|t) & = -z - (b-d-z) \phi(z,T|t) + b \phi(z,T|t)^2
    \end{split}
\end{equation}
backwards in time starting from $\phi(z,T|T)=0$ at $t=T$.
While $\phi(z,T|t)$ is small, this is approximately solved by
\begin{equation}
\begin{split}
    \phi(z,T|t) & = z e^{\int_t^T (b-d-z)dt'} \int_t^T e^{-\int_\tau^T (b-d-z)dt'} d\tau \\
    & = z e^{(b-d-z)(T-t)} \int_t^T e^{-(b-d-z)(T-\tau)} d\tau \\
    & = z e^{(b-d-z)(T-t)}\left[1   - e^{-(b-d-z)(T-t)}\right]/(b-d-z) \\
    & = \frac{z}{b-d-z}\left[e^{(b-d-z)(T-t)} - 1\right] = \frac{z}{\gamma_0 - s - z}\left[e^{\gamma_0 (T-t) - (s+z) (T-t)} - 1\right]
\end{split}
\end{equation}
where $\gamma_0$ is the growth rate of the background population.

At longer times when $z e^{\gamma_0 (T-t)}\sim 1$ and $\phi$ is no longer small, $\phi$ tends towards a constant value determined by the same quadratic equation as above.
This limit is neither interesting or relevant for the present purpose, since there are very few lineages that emerged early enough to have saturated $\phi$.
Instead, we need to average $\phi$ (the linear approximation) over all the time points when the lineage could have arisen.
\begin{equation}
    \begin{split}
        \langle \phi \rangle &\sim \int_t^T dt' \; e^{-\gamma_0(T-t')} \frac{z (e^{\gamma_0(T-t') - (s+z)(T-t')}-1)}{(\gamma_0 - s - z)}\\
        &= \int_t^T dt' \;  \frac{z (e^{-(s+z)(T-t')}-e^{-\gamma_0(T-t')})}{(\gamma_0 - s - z)}\\
        &\approx \begin{cases}
            \frac{z}{\gamma_0-s-z} [ \frac{1}{z+s} - \frac{1}{\gamma_0}] & s(T-t)\gg 1\\
            \frac{z}{\gamma_0-s-z} [ (T-t) - \frac{(s+z)(T-t)^2}{2} - \frac{1}{\gamma_0}] & s(T-t)<1 \\
        \end{cases}
    \end{split}
\end{equation}
This derivation assumed that $\gamma_0(T-t)\gg1$, i.e.~that the overall population size has expanded substantially.
The most relevant fitness effects will be those with $s(T-t)>1$ but $\gamma_0>s$ such that the mutant lineage has been amplified but its fitness difference to the background is a perturbation.
In this case, the above simplifies to
\begin{equation}
    \langle \phi \rangle \approx \frac{z}{\gamma_0(z+s)}
\end{equation}

In a variant that has been growing with rate $\gamma_0$ for a time $\tau=T-t$ and sampled with $z=\epsilon$, we thus expect that the number of times we observe separate mutant lineages depends on $s$ as
\begin{equation}
    \langle \phi \rangle \approx \frac{\epsilon}{\gamma_0(\epsilon+s)}
\end{equation}
This has a very similar behavior as the solution for constant population size, which suggests that the overall dependence on $s$ is robust and we can assume that the number of times a mutation is observed is inversely proportional to its effect on fitness.
In a constant population, this relationship breaks down for dense sampling $\epsilon > \sqrt{s}$.
In growing population, the approximation fails if the product of fitness effect and the time over which the variant has grown, $s\tau$, is large, i.e., if the fitness cost is substantial on this time scale.
Outside this parameter range, there is still a dependence on $s$, but it is weaker.

\section{Supplementary figures}
